Towards a standardized evaluation of imputation methodology

### 2022-03-31 Manuscript/Towards_a_standardized_evaluation_of_imputation_methodology.tex
Submitted version

\begin{document}
%\DOIsuffix{bimj.DOIsuffix}
\DOIsuffix{bimj.200100000}
\Volume{52}
\Issue{61}
\Year{2010}
\pagespan{1}{}
\keywords{Evaluation; Imputation; Missing data; Simulation studies;\\ [1pc]
% \noindent \hspace*{-4.2pc} Supporting Information for this article is available from Github, \break \hspace*{-4pc} \underline{github.com/gerkovink/StandardizedEvaluation}.
% } %%% semicolon and fullpoint added here for keyword style
\noindent \hspace*{-4pc} Supporting Information for this article is available from \underline{gerkovink.com/evaluation}.\\
% \hspace*{-4pc} \underline{gerkovink.com/evaluation}. 
}  %%% semicolon and fullpoint added here for keyword style

@@

\begin{abstract}
Developing new imputation methodology has become a very active field. Unfortunately, there is no consensus on how to perform simulation studies to evaluate the properties of imputation methods. In this paper, we propose a move towards a standardized evaluation of imputation methods. To demonstrate the need for standardization, we highlight a set of potential pitfalls that bring forth a chain of potential problems in the objective assessment of the performance of imputation routines. This may lead to sub-optimal use of multiple imputation in practice. Additionally, we suggest a course of action for simulating and evaluating missing data problems.
\end{abstract}

@@

To evaluate the ability of an imputation routine to handle missingness, a form of truth has to be established. Those who perform simulation studies are in the luxury position to establish the truth beforehand by choosing a data-generating mechanism. Data-generating mechanisms define how a complete dataset is obtained at the start of each simulation repetition. There are two general approaches to generating complete data: (i) model-based simulation, in which data are drawn from a known statistical model or probability distribution, such as the multivariate normal distribution; and (ii) design-based simulation, where data are sampled (with replacement) from a sufficiently large observed set, such as from official registers.

The problem with model-based data-generating mechanisms is that a method's performance on simulated data may not translate to empirical data. Real-life data hardly ever follow a given theoretical distribution, so there is no guarantee that simulation results are generalizable. Moreover, data are often generated such that the problem that is being studied is most pronounced, e.g. with consistently high correlations between groups of variables. This results in simulated data that contain such valuable information structures that, no matter what type of missingness would subsequently be induced, the observed parts of the data will still hold much (if not all) of the information about the missing part. Unsurprisingly, the performance of any imputation method will then be evaluated as good. Another threat to the generalizability of model-based simulations is the use of a single model for both data generation and imputation. If data are generated following a model that is also used for imputing the data, the imputation approach will be deemed good (or better than other methods) purely due to the evaluated conditions being in favor of the problem that is studied. Other (unfair) comparative advantages in favor of a certain imputation method may occur due to characteristics of the generated data, such as the number of observations, the number of variables, the variable type(s), and the coherence between variables. In contrast to design-based studies, such characteristics are not always explicit simulation conditions, which may give a false sense of objectivity.

An obvious problem with design-based simulation is that obtaining a large dataset without missing entries can be very challenging. Most real-world data contains at least some missing entries, for which the true underlying missing data model is--by definition--unknown. Therefore, the simulator needs to deal with missingness in \textit{some} way before incomplete empirical data can serve as comparative truth in the simulations. It may seem like an intuitive solution to only draw complete cases from the large dataset, which would indeed yield complete samples. However, there may be inherent differences between cases with and cases without any missing values, due to the unknown missing data model. Only sampling complete cases from the data set may thus result in samples that fail to capture all relevant real-world conditions, which in turn refutes the main reason for using design-based simulation. Another way to deal with missingness in a design-based simulation is to impute the incomplete dataset once, to obtain a single completed dataset to draw samples from. Unfortunately, this practice may favor the imputation method that was used in this initial imputation step throughout any further evaluations. Just to be clear: leaving the missingness as-is and inducing additional missing values to impute could not work here, because we wouldn't have a real and unbiased comparative truth.

@@

\begin{table}[tb]
\begin{center}
\caption{Missingness mechanisms.}
\label{table:mech}
\begin{tabular}{lll}
\hline
      & Interpretation and consequences of the mechanism \\
\hline  
MCAR  & Missing completely at random: The probability to be missing is the same for all \\
      & cases. In other words, the missing values are missing at random and the observed \\
      & values are observed at random. \\
MAR   & Missing at random: The probability to be missing is the same within groups of \\
      & cases defined by the observed data only. The essence is that observed relations in \\
      & the data inform the missingness, such that MCAR may be assumed within the \\ 
      & observed groups. \\
MNAR  & Missing not at random: The probability to be missing depends on unobserved \\
      & aspects of the data. The cause of the missingness is unknown and cannot be \\ 
      & inferred from the observed data. This is considered non-ignorable missingness \\
      & \citep[see e.g.][]{rubi76}. \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

Missingness mechanisms are usually assumed to be random (MAR; see Table \ref{table:mech} for definitions). However, the missingness mechanisms that are induced in simulation studies do not necessarily match any true missingness generating mechanisms in incomplete empirical data, nor the mechanisms that are typically assumed when imputing such data. On the one hand, there is MCAR missingness, which may be unlikely as the true missingness generating mechanism, but would yield insightful evaluations in a simulation setting. On the other hand, there is MNAR, which is arguably a quite reasonable source of many real-world missingness scenarios but is generally ignored except for those evaluations that are specifically targeted at non-ignorable applications. A disconnect between induced missingness, real missingness, and assumed missingness may result in simulation studies that are not as informative as they could be.

@@

Although MAR missingness is often considered as a simulation condition, the problem of spurious MAR is generally overlooked. With MAR missingness mechanisms, observed relations in the data are used to induce missingness during simulation (e.g. weight is made incomplete based on observed gender to emulate a situation wherein one gender is less likely to disclose their weight). These relations in the observed data may, however, be weak or non-existent. If MAR is induced based on weaker relations in the data, claims for a method's applicability to situations where the missingness is random become less valid. The most extreme example would be when MAR is induced from data without multivariate relations. The inferential implications of the missingness would then mimic those of MCAR, even though the missingness is random. In fact, this would amount to one of the special cases under which complete case analysis would be more efficient than multiple imputation \citep[see e.g.][\S 2.7]{buur18}: the missingness does not depend on the incomplete variable. This property might be useful in practice, but considering it as a condition to evaluate performance under MAR missingness is pointless.

@@

\begin{table}[tb]
\begin{center}
\caption{Steps to consider in imputation simulation studies.}
\label{table:steps}
\begin{tabular}{lll}
\hline
                        & Simulation aspects under consideration \\
\hline  
1. Set scope            & Aim(s), missing data method(s) to evaluate, simulation design \\
2. Obtain truth         & Data-generating mechanism(s), estimand(s), sampling variance \\
3. Induce missingness   & Missingness mechanism(s), missing data pattern(s) \\
4. Apply methods        & Imputation method(s), analytic method(s) \\
5. Evaluate imputations & Algorithmic convergence, distributional characteristics, imputation \\
                        & model fit \\
6. Evaluate performance & Performance measures, qualitative judgment \\
7. Report               & Text, visualization, checklist, simulation script \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

Next, missing data should be induced conform a model that is dependent on the observed data (i.e. a MAR mechanism). Arguably, MAR is the most often-assumed mechanism in practice, and should thus be evaluated carefully. A straightforward technique for inducing univariate MAR missingness is described in \citet[][\S 3.2.4]{buur18}, generalizations to multivariate MAR missingness can be found in \citet{ampute}. If the missingness is to be induced in longitudinal data autoregressive MAR models can be useful \citep[see e.g.][model 2 and model 3]{shara2015randomly} can be useful. It is advisable to investigate varying shapes of MAR missingness to achieve a more realistic indication of the robustness of the imputation performance across the range of random missingness. The effects of different types of MAR mechanisms are described in \citet{scho18}.

@@

\begin{table}[htb]
\begin{center}
\caption{Suggested univariate missingness proportions.}
\label{table:prop}
\begin{tabular}{lll}
\hline
      & Reasoning for suggested missingness proportion \\
\hline  
10\%  & Depending on the size of the data, this percentage can be considered as a lower \\
      & bound of realistic evaluation. Anything less than 10\% may be of little influence on \\
      & the true data inference. Performance of a missing data method should at least be \\
      & acceptable for most missing data problems. \\ % & \\
25\%  & This is a fair amount of missingness and will, depending on the observed data \\
      & information, have a noticeable influence on the completed data inference. When \\
      & compared to the condition with 10\% missingness, the inference obtained under \\
      & 25\% missingness should be less certain (i.e. confidence/credibility interval width \\
      & should increase), but estimates should still be properly covered and the statistical \\
      & properties of the missing data method should be sound. In practice, at least to our \\
      & at least to our experience in social sciences and official statistics, 25\% \\
      & univariate missingness can easily be considered as a realistic missingness percentage. \\ % & \\
50\%  & Performance under 50\% percent simulated missingness will most likely be impacted \\ 
      & severely. Depending on how the missingness mechanism interacts with the simulated \\
      & data, some imputation techniques may yield estimates that are under-covered such that \\
      & the completed data inference should not be deemed valid anymore. If a method yields \\
      & acceptable inference under 50\% MAR missingness, we can determine that the statistical \\
      & properties of the imputation methodology are sound. \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

Imputation method performance should be quantified using appropriate measures. It depends on the specifics of each simulation study which performance measures would be most suitable. In general, we would say that each multiple imputation routine should be evaluated on at least the following points:

@@

After the careful consideration and execution of all simulation aspects, the evaluations should be properly reported. We encourage simulators to document their choices and to be explicit in their descriptions. For example, the simulation design may be presented textually, in a flow chart or as a block of pseudo-code, whereas missingness mechanisms may be written as a function of the data or displayed graphically. Ideally, this should be supplemented by an online repository with all of the data and code required to reproduce the simulation results. To aid simulators in reporting and move towards standardization in evaluation, we provide a draft version for reporting guidelines in Appendix A.1 (also available from \underline{gerkovink.com/evaluation}). We invite the readers of this paper to contribute to its development.

@@

\begin{table}[ht]
\caption{Checklist for reporting on imputation methodology evaluations}
\label{table:check}
\begin{tabular}{ll}
% \hline
%   & Suggested aspects to report                                                       \\
\hline
1 & Simulation scope                                                                   \\ \cline{2-2}
  & Aim                                                                                \\
  & Design (incl. pseudo-code or flow diagram)                                         \\
  & Number of simulation repetitions                                                   \\ \hline
2 & Comparative truth                                                                  \\ \cline{2-2}
  & Data-generating mechanism (model-based or design-based)                            \\
  & Sampling variance                                                                  \\
%  & Data characteristics (number of observations, number of variables, variable types, \\
%  & coherence between variables incl. data structures e.g. clustering)                 \\
  & Data characteristics (incl. multivariate relations and structures e.g. clustering) \\
  & Estimand                                                                           \\ \hline 
3 & Induced missingness                                                                \\ \cline{2-2}
  & Missingness mechanism (incl. type or functional form of the missing data   model)  \\
  & Missingness pattern (incl. missingness proportion)                                 \\ \hline 
4 & Applied methods                                                                    \\ \cline{2-2}
  & Imputation methods (incl. parameters e.g. the number of imputations)               \\
  & Analytic methods (incl. calculation of standard errors e.g. pooling   rules)       \\
  & Reference method (e.g. complete case analysis)                                    \\ \hline 
5 & Imputation evaluation                                                              \\ \cline{2-2}
  & Imputation-generating process (e.g. algorithmic non-convergence)                   \\
  & Imputation model fit (e.g. posterior predictive checks)                            \\
  & Distributional characteristics (e.g. plausibility of imputed values)               \\ \hline 
6 & Performance evaluation                                                             \\ \cline{2-2}
  & Statistical properties (e.g. confidence validity)                                  \\
  & Comparative performance (e.g. predictive accuracy)                                 \\ \hline
\end{tabular}
\end{table}

>>>>

\begin{document}
%\DOIsuffix{bimj.DOIsuffix}
\DOIsuffix{bimj.200100000}
\Volume{52}
\Issue{61}
\Year{2010}
\pagespan{1}{}
\keywords{Evaluation; Imputation; Missing data; Simulation studies;\\ [1pc]
% \noindent \hspace*{-4.2pc} Supporting Information for this article is available from Github, \break \hspace*{-4pc} \underline{github.com/gerkovink/StandardizedEvaluation}.
% } %%% semicolon and fullpoint added here for keyword style
\noindent \hspace*{-4pc} Supporting Information for this article is available from \underline{www.gerkovink.com/evaluation}.\\
% \hspace*{-4pc} \underline{gerkovink.com/evaluation}. 
}  %%% semicolon and fullpoint added here for keyword style

@@

\begin{abstract}
Developing new imputation methodology has become a very active field. Unfortunately, there is no consensus on how to perform simulation studies to evaluate the properties of imputation methods. In this paper, we propose a move towards a standardized evaluation of imputation methodology. To demonstrate the need for standardization, we highlight a set of potential pitfalls that bring forth a chain of potential problems in the objective assessment of the performance of imputation routines. This may lead to sub-optimal use of  imputation in practice. Additionally, we suggest a course of action for simulating and evaluating missing data problems.
\end{abstract}

@@

To evaluate the ability of an imputation routine to handle missingness, a form of truth has to be established. Those who perform simulation studies are in the luxury position to establish the truth beforehand by choosing a data-generating mechanism. Data-generating mechanisms define how a complete dataset is obtained at the start of each simulation repetition. There are two general approaches to generating complete data: (i) model-based simulation, in which data are drawn from a known statistical model or probability distribution, such as the multivariate normal distribution; and (ii) design-based simulation, where data are sampled (with replacement) from a sufficiently large observed set, such as official registers.

The problem with model-based data-generating mechanisms is that a method's performance on simulated data may not translate to empirical data. Real-life data hardly ever follow a given theoretical distribution, so there is no guarantee that simulation results are generalizable. Moreover, data are often generated such that the problem being studied is most pronounced, e.g. with consistently high correlations between groups of variables. This results in simulated data that contain such valuable information structures that, no matter what type of missingness would subsequently be induced, the observed parts of the data will still hold much (if not all) of the information about the missing part. Unsurprisingly, the performance of any imputation method will then be evaluated as good. Another threat to the generalizability of model-based simulations is the use of a single model for both data generation and imputation. If data are generated following a model that is also used for imputing the data, the imputation approach will be deemed good (or better than other methods) purely due to the evaluated conditions being in favor of the problem that is studied. Other (unfair) comparative advantages in favor of a certain imputation method may occur due to characteristics of the generated data, such as the number of observations, the number of variables, the variable type(s), and the coherence between variables. In contrast to design-based studies, such characteristics are not always explicit simulation conditions, which may give a false sense of objectivity.

An obvious problem with design-based simulation is that obtaining a large dataset without missing entries can be very challenging. Most real-world data contains at least some missing entries, for which the true underlying missing data model is--by definition--unknown. Therefore, the simulator needs to deal with missingness in \textit{some} way before incomplete empirical data can serve as comparative truth in the simulations. It may seem like an intuitive solution to only draw complete cases from the large dataset, which would indeed yield complete samples. However, there may be inherent differences between cases with and cases without any missing values, due to the unknown missing data model. Only sampling complete cases from the data set may thus result in samples that fail to capture all relevant real-world conditions, which in turn refutes the main reason for using design-based simulation. Another way to deal with missingness in a design-based simulation is to impute the incomplete dataset once, to obtain a single completed dataset to draw samples from. Unfortunately, this practice may favor the imputation method that was used in this initial imputation step throughout any further evaluations. Just to be clear: leaving the missingness as-is and inducing additional missing values to impute is no option here, because we would not have a real and unbiased comparative truth.

@@

\begin{table}[tb]
\begin{center}
\caption{Missingness mechanisms.}
\label{table:mech}
\begin{tabular}{lll}
\hline
      & Interpretation and consequences of the mechanism \\
\hline  
MCAR  & Missing completely at random: The probability to be missing is the same for all \\
      & cases. In other words, the missing values are missing at random and the observed \\
      & values are observed at random. \\ \noalign{\smallskip}
MAR   & Missing at random: The probability to be missing is the same within groups of \\
      & cases defined by the observed data only. The essence is that observed relations in \\
      & the data inform the missingness, such that MCAR may be assumed within the \\ 
      & observed groups. \\ \noalign{\smallskip}
MNAR  & Missing not at random: The probability to be missing depends on unobserved \\
      & aspects of the data. The cause of the missingness is unknown and cannot be \\ 
      & inferred from the observed data. This is considered non-ignorable missingness \\
      & \citep[see e.g.][]{rubi76}. \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

Missingness mechanisms are usually assumed to be random (MAR; see Table \ref{table:mech} for definitions). However, the missingness mechanisms that are induced in simulation studies do not necessarily match any true missingness generating mechanisms in incomplete empirical data, nor the mechanisms that are typically assumed when imputing such data. On the one hand, there is MCAR missingness, which is unlikely to be the true missingness generating mechanism, but would yield insightful evaluations in a simulation setting. On the other hand, there is MNAR, which is arguably a quite reasonable source of many real-world missingness scenarios but is generally ignored except for those evaluations that are specifically targeted at non-ignorable applications. A disconnect between induced missingness, real missingness, and assumed missingness may result in simulation studies that are not as informative as they could be.

@@

Although MAR missingness is often considered as a simulation condition, the problem of spurious MAR is generally overlooked. With MAR missingness mechanisms, observed relations in the data are used to induce missingness during simulation (e.g. weight is made incomplete based on observed gender to emulate a situation wherein one gender is less likely to disclose their weight). These relations in the observed data may, however, be weak or non-existent. If MAR is induced based on weaker relations in the data, claims for a method's applicability to situations where the missingness is random become less valid. The most extreme example would be when MAR is induced from data without multivariate relations. The inferential implications of the missingness would then mimic those of MCAR, even though the missingness is random. In fact, this would amount to one of the special cases under which complete case analysis would be more efficient than imputation \citep[see e.g.][\S 2.7]{buur18}: the missingness does not depend on the incomplete variable. This property might be useful in practice, but considering it as a condition to evaluate performance under MAR missingness is pointless.

@@

\begin{table}[tb]
\begin{center}
\caption{Steps to consider in imputation simulation studies.}
\label{table:steps}
\begin{tabular}{lll}
\hline
                        & Simulation aspects under consideration \\
\hline  
1. Set scope            & Aim(s), missing data method(s) to evaluate, simulation design \\
2. Obtain truth         & Data-generating mechanism(s), estimand(s), sampling variance \\
3. Induce missingness   & Missingness mechanism(s), missing data pattern(s) \\
4. Apply methods        & Imputation method(s), analytic method(s) \\
5. Evaluate imputations & Imputation-generating process, imputation model fit \\
6. Evaluate performance & Performance measures, qualitative judgment \\
7. Report               & Text, visualization, checklist, simulation script \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

Next, missing data should be induced conform a model that is dependent on the observed data (i.e. a MAR mechanism). Arguably, MAR is the most often-assumed mechanism in practice, and should thus be evaluated carefully. A straightforward technique for inducing univariate MAR missingness is described in \citet[][\S 3.2.4]{buur18}, generalizations to multivariate MAR missingness can be found in \citet{ampute}. If the missingness is to be induced in longitudinal data autoregressive MAR models can be useful \citep[see e.g.][model 2 and model 3]{shara2015randomly}. It is advisable to investigate varying shapes of MAR missingness to achieve a more realistic indication of the robustness of the imputation performance across the range of random missingness. The effects of different types of MAR mechanisms are described in \citet{scho18}.

@@

\begin{table}[htb]
\begin{center}
\caption{Suggested univariate missingness proportions.}
\label{table:prop}
\begin{tabular}{lll}
\hline
      & Reasoning for suggested missingness proportion \\
\hline  
10\%  & Depending on the size of the data, this percentage can be considered as a lower \\
      & bound of realistic evaluation. Anything less than 10\% may be of little influence on \\
      & the true data inference. Performance of a missing data method should at least be \\
      & acceptable for most missing data problems. \\ \noalign{\smallskip} % & \\
25\%  & This is a fair amount of missingness and will, depending on the observed data \\
      & information, have a noticeable influence on the completed data inference. When \\
      & compared to the condition with 10\% missingness, the inference obtained under \\
      & 25\% missingness should be less certain (i.e. confidence/credibility interval width \\
      & should increase), but estimates should still be properly covered and the statistical \\
      & properties of the missing data method should be sound. In practice, at least to our \\
      & at least to our experience in social sciences and official statistics, 25\% \\
      & univariate missingness can easily be considered as a realistic missingness percentage. \\ \noalign{\smallskip} % & \\
50\%  & Performance under 50\% percent simulated missingness will most likely be impacted \\ 
      & severely. Depending on how the missingness mechanism interacts with the simulated \\
      & data, some imputation techniques may yield estimates that are under-covered such that \\
      & the completed data inference should not be deemed valid anymore. If a method yields \\
      & acceptable inference under 50\% MAR missingness, we can determine that the statistical \\
      & properties of the imputation methodology are sound. \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

Imputation method performance should be quantified using appropriate measures. It depends on the specifics of each simulation study which performance measures would be most suitable. In general, we recommend to evaluate at least the following points:

@@

After the careful consideration and execution of all simulation aspects, the evaluations should be properly reported. We encourage simulators to document their choices and to be explicit in their descriptions. For example, the simulation design may be presented textually, in a flow chart or as a block of pseudo-code, whereas missingness mechanisms may be written as a function of the data or displayed graphically. Ideally, this should be supplemented by an online repository with all of the data and code required to reproduce the simulation results. To aid simulators in reporting and move towards standardization in evaluation, we provide a draft version for reporting guidelines in Appendix A.1 (also available from \underline{www.gerkovink.com/evaluation}). We invite the readers of this paper to contribute to its development.

@@

\begin{table}[ht]
\caption{Checklist for reporting on imputation methodology evaluations.}
\label{table:check}
\begin{tabular}{ll}
% \hline
%   & Suggested aspects to report                                                       \\
\hline
1 & Simulation scope                                                                   \\ \cline{2-2}
  & Aim                                                                                \\
  & Design (incl. pseudo-code or flow diagram)                                         \\
  & Number of simulation repetitions                                                   \\ \hline
2 & Comparative truth                                                                  \\ \cline{2-2}
  & Data-generating mechanism (model-based or design-based)                            \\
  & Sampling variance                                                                  \\
%  & Data characteristics (number of observations, number of variables, variable types, \\
%  & coherence between variables incl. data structures e.g. clustering)                 \\
  & Data characteristics (incl. multivariate relations and structures e.g. clustering) \\
  & Estimand                                                                           \\ \hline 
3 & Induced missingness                                                                \\ \cline{2-2}
  & Missingness mechanism (incl. type or functional form of the missing data   model)  \\
  & Missingness pattern (incl. missingness proportion)                                 \\ \hline 
4 & Applied methods                                                                    \\ \cline{2-2}
  & Imputation methods (incl. parameters e.g. the number of imputations)               \\
  & Analytic methods (incl. calculation of standard errors e.g. pooling   rules)       \\
  & Reference method (e.g. complete case analysis)                                    \\ \hline 
5 & Imputation evaluation                                                              \\ \cline{2-2}
  & Imputation-generating process (e.g. algorithmic non-convergence)                   \\
  & Imputation model fit (e.g. posterior predictive checks)                            \\
  & Distributional characteristics (e.g. plausibility of imputed values)               \\ \hline 
6 & Performance evaluation                                                             \\ \cline{2-2}
  & Statistical properties (e.g. confidence validity)                                  \\
  & Comparative performance (e.g. predictive accuracy)                                 \\ \hline
\end{tabular}
\end{table}
